Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2022-10-07 analysis/paper/paper.qmd
Update paper.qmd

```{r}

#| label: fig-sp-hand-made
#| fig-cap: Stemmed points from Yongsandong site. 1-6: plain stemmed points, 7-8: one side denticulate stemmed points.

@@

Though the stemmed points are generally assumed as hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to understand the function of the tools without knowing their complete shape attached to other components. Preserved wooden components of projectile tools are too few and rare to generalize the overall scale and variability of the tools used in prehistory [@shea2006origins]. -@Lee_Kong_2002 claimed to name the stemmed points as 'stemmed tools' because of the diverse shape of stemmed points, which could represent different functions as well as the uncertainty of complete tool shape. Other researchers agree that some stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called as stemmed points but stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] (@fig-sp-hand-made: 7-8).

@@

After the first evidence of the Korean Late Paleolithic discovered at Seokjangri site including blade and a stemmed base, around 300 stemmed points have been found in Korea [@Sohn_1967; @lee2019were]. Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we gathered those that retained the widest and thickest part of specimens. Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but accompanied with other Late Paleoltihic artifacts. Those criteria resulted in a sample of 173 stemmed points from 36 assemblages unearthed from 29 sites spanning the period 44-10 ka (Figure 2). The dimensions of the 173 stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of Korean local museums and archaeological institutions.

@@

```{r}

#| label: fig-map
#| fig-cap: Korean Paleolithic sites mentioned in this study
#| fig-width:   4.45
#| fig-height:  4.5

@@

map <-
  get_stamenmap(bbox = c(left = 124,
                         bottom = 33,
                         right = 	130,
                         top = 39),
                zoom = 10)

@@

```{r}
#| label: fig-tcsa-all-sp
#| fig-cap: Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes show TSCA ranges for various weapon types.
#| fig-width:   4.5
#| fig-height:  4.4

@@

```{r}

#| label: fig-tcsa-raw-materials
#| fig-cap: Histograms of TCSA values by lithic raw material.
#| fig-width:   4.5
#| fig-height:  6

@@

```{r}

#| label: tcsa-sites
#| fig-cap: Histograms of TCSA values by archaeological site.
#| fig-width:   4.5
#| fig-height:  6

@@

We computed TCSA for sites that contain more than 5 stemmed points to observe the pattern. As shown in @tcsa-sites, the sites with fewer stemmed points are grouped under the category of 'other'.  Among the selected sites, Suyanggae has the most stemmed points and the widest range of TCSA. We can assume that people made stemmed points for a variety of different purposes in Suyanggae. The TCSA values for Nosanri and Sibuk are skewed, which indicates less diverse functions for stemmed points, and lower. Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but exhibit different distribution of TCSA values. Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.

@@

```{r}

#| label: fig-tcsa-size
#| fig-cap: Artifact size and TSCA values by lithic raw material
#| fig-width:   4.5
#| fig-height:  4

@@

```{r}

#| label: artifact-size-cluster
#| fig-cap: A. Distribution of artefact weight showing three clusters. B. TCSA values for all artefacts with size classes indicated by data point shape.
#| fig-width:   4.5
#| fig-height:  4

@@

We then conducted univariate cluster analysis of stemmed points by weight and it revealed three groups of clusters (A of @artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`). Cluster 1, the smallest (lightest) artifacts, is the lower TCSA, compared to Cluster 2. TCSA of Cluster 3 is the highest, except for one artifact, which is lower than 50 (A of @artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`). There are a few overlaps, but we can still assume that artifacts in different clusters might be made for different purposes.

@@

```{r}

#| label: tcsa-radio-carbon
#| fig-cap: Distribution of TCSA values by time
#| fig-width:   4.5
#| fig-height:  4

@@

 ggplot(tcsa_site_ass, aes(x = min_tcsa, y = reorder(label_site, age_ka), color = label_site)) +
   geom_point() +
   theme(legend.position="none") +
   labs(x = "TCSA", 
       y = "Assemblage")

```
Using radiocarbon dates, we arranged 24 assemblages in chronological order (@tcsa-radio-carbon). Our analysis shows that early assemblages (i.e. 45-24 ka) tend to possess numerous stemmed points such as two from the Suyanggae site and they have a wide range of TCSA, indicating multiple usages of stemmed points. Overall, TCSA from assemblages that have more than 3 stemmed points vary considerably, which makes it difficult to define a single function of the artifacts. There is also no clear temporal pattern in assemblages with a fewer number of artifacts.

>>>>

```{r}
#| label: fig-sp-hand-made
#| fig-cap: "Stemmed points from Yongsandong site. 1-6: plain stemmed points, 7-8: one side denticulate stemmed points."

@@

Though the stemmed points are generally assumed as hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to understand the function of the tools without knowing their complete shape attached to other components. Preserved wooden components of projectile tools are too few and rare to generalize the overall scale and variability of the tools used in prehistory [@shea2006origins]. -@Lee_Kong_2002 claimed to name the stemmed points as 'stemmed tools' because of the diverse shape of stemmed points, which could represent different functions as well as the uncertainty of complete tool shape. Other researchers agree that some stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called as stemmed points but stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.

@@

After the first evidence of the Korean Late Paleolithic discovered at Seokjangri site including blade and a stemmed base, around 300 stemmed points have been found in Korea [@Sohn_1967; @lee2019were]. Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we gathered those that retained the widest and thickest part of specimens. Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but accompanied with other Late Paleoltihic artifacts. Those criteria resulted in a sample of 173 stemmed points from 36 assemblages unearthed from 29 sites spanning the period 44-10 ka (@fig-map). The dimensions of the 173 stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of Korean local museums and archaeological institutions.

@@

```{r}
#| label: fig-map
#| fig-cap: "Korean Paleolithic sites mentioned in this study"
#| fig-width:   4.45
#| fig-height:  4.5

@@

# if we have already downloaded the map tiles, skip doing it again
if (file.exists(here::here("analysis/data/raw_data/map.RData"))) {

  # load the the map tiles have already been downloaded
  load(file = here::here("analysis/data/raw_data/map.RData"))

  } else {

    # download map tiles for our study area if we don't 
    # have them already
map <-
  get_stamenmap(bbox = c(left = 124,
                         bottom = 33,
                         right = 	130,
                         top = 39),
                zoom = 10)
# and save the tiles so we don't have to do this every time we render
save(map, file = here::here("analysis/data/raw_data/map.RData"))

}

@@

```{r}
#| label: fig-tcsa-all-sp
#| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes show TSCA ranges for various weapon types."
#| fig-width:   4.5
#| fig-height:  4.4

@@

```{r}
#| label: fig-tcsa-raw-materials
#| fig-cap: "Histograms of TCSA values by lithic raw material."
#| fig-width:   4.5
#| fig-height:  6

@@

```{r}
#| label: fig-tcsa-sites
#| fig-cap: "Histograms of TCSA values by archaeological site."
#| fig-width:   4.5
#| fig-height:  6

@@

We computed TCSA for sites that contain more than 5 stemmed points to observe the pattern. As shown in @fig-tcsa-sites, the sites with fewer stemmed points are grouped under the category of 'other'.  Among the selected sites, Suyanggae has the most stemmed points and the widest range of TCSA. We can assume that people made stemmed points for a variety of different purposes in Suyanggae. The TCSA values for Nosanri and Sibuk are skewed, which indicates less diverse functions for stemmed points, and lower. Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but exhibit different distribution of TCSA values. Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.

@@

```{r}
#| label: fig-tcsa-size
#| fig-cap: "Artifact size and TSCA values by lithic raw material"
#| fig-width:   4.5
#| fig-height:  4

@@

```{r}
#| label: fig-artifact-size-cluster
#| fig-cap: "A. Distribution of artefact weight showing three clusters. B. TCSA values for all artefacts with size classes indicated by data point shape."
#| fig-width:   4.5
#| fig-height:  4

@@

We then conducted univariate cluster analysis of stemmed points by weight and it revealed three groups of clusters (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`). Cluster 1, the smallest (lightest) artifacts, is the lower TCSA, compared to Cluster 2. TCSA of Cluster 3 is the highest, except for one artifact, which is lower than 50 (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`). There are a few overlaps, but we can still assume that artifacts in different clusters might be made for different purposes.

@@

```{r}
#| label: fig-tcsa-radio-carbon
#| fig-cap: "Distribution of TCSA values by time"
#| fig-width:   4.5
#| fig-height:  4

@@

 ggplot(tcsa_site_ass, aes(x = min_tcsa, y = reorder(label_site, age_ka), color = label_site)) +
   geom_point() +
   theme_minimal() +
   theme(legend.position="none") +
   labs(x = "TCSA", 
       y = "Assemblage")

Using radiocarbon dates, we arranged 24 assemblages in chronological order (@fig-tcsa-radio-carbon). Our analysis shows that early assemblages (i.e. 45-24 ka) tend to possess numerous stemmed points such as two from the Suyanggae site and they have a wide range of TCSA, indicating multiple usages of stemmed points. Overall, TCSA from assemblages that have more than 3 stemmed points vary considerably, which makes it difficult to define a single function of the artifacts. There is also no clear temporal pattern in assemblages with a fewer number of artifacts.
